A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Response-to-Reviewers-1.Rmd

---
title: "Response to Reviewers"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Editor:

Thank you for your decision letter of May 5th, concerning our jointly authored submission *A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain*.

We would like to express our gratitude to the three anonymous reviewers for their thoughtful comments, which we addressed in this revision.

In what follows, we describe the actions taken in response to the feedback received. We trust that you, and the reviewers, will find that we have addressed all recommendations in a thorough and complete manner.

We look forward to your opinion regarding the revisions of this paper.

Sincerely,

The authors

\newpage

## Reviewer 1

Narrative

This timely paper examines the association between environmental variables and COVID-19 incidence in Spain.  It is well-written and contains a robust statistical analysis. The SUR method is very robust and appropriate to address the authors’ research questions; while it is very well articulated in the text.  The results and discussion are well presented and the exhibited decrease in Rho (spatial autocorrelation) over time is particularly very insightful.

I believe that the paper is suitable for publication after some minor revisions.

> *Thank you for your kind comments and positive assessment of the paper.*

Please see my specific suggestions and comments below:

Introduction and Purpose

Page 1, Lines 42-43: “the SARS-CoV2 virus has threatened to overrun health systems the world over” – sentence is awkward and should be rephrased to “around the world” or “global health systems”.

Page 1, Line 47: “arrested the spread” should be “mitigated the spread of the virus”.

Page 1, Line 48: “helped to keep a bad situation from becoming even worse” is awkward and should be rephrased.  Do you have any citations regarding the efficacy of social distancing?

> *These items were edited as per your suggestion, and we have included references as requested.*

Page 2, Lines 24-28: You need to add the degree symbol ° in front of C.  Please do so throughout the paper.

> *Done.*

Page 3, Lines 32-42: This paragraph does not really belong in the introduction.  You are discussing the key findings and results here, which belong in the discussion and conclusion.  Instead, I suggest adding some hypotheses and research questions.

> *The introduction was reworked and most of this material was moved to the Background. This helped to make the introduction shorter and more engaging.*

Page 3, Line 47: The link should be a footnote, rather than in the main body of the text.

> *Changed to a footnote.*

Context and Data Methods

Page 3, Line 51: “January 31th” should be “January 31st, 2020”.

> *Corrected.*

Page 3, Lines 9-10: Is the relatively slow start due to testing lags? You may want to add something to clarify.  In the United States, many people who died or were infected with the virus in January and February were considered influenza cases and not COVID-19 cases.

> *This is true. We have edited the text to note the relationship with testing capabilities.*

Page 4, Lines 13-14: Was Madrid the original hotspot of COVID-19 because of testing resources and population?

> *The original hotspot was actually Álava in Pais Vasco; after reorganizing the sections, we hope that it is clearer that our hypothesis is that population and GDP are related to the early flare of the outbreak in Álava and Madrid (the first and second wealthiest Autonomous Communities in Spain, in GDP per capita).*

Section 2.2: This is all great information, but it almost reads like a literature review and could be condensed and discussed in the introduction.  I think the paper can be rearranged to Introduction and Context; then Data and Methods.

> *The sections have been reorganized following this suggestion. The introduction is now a little bit shorter and more in-depth discussion of the state of knowledge is now in the Background.*

Page 6, Lines 29-30: US should be abbreviated as U.S.  Please fix throughout paper.

> *Done.*

Table 1: Is the sunshine variable missing from the table?

> *It was. We have corrected this oversight.*

Methods

Page 8, Line 15. “varaibles” is misspelled. Should be “variables”.

> *Fixed.*

Is the model actually a space-time model, rather than spatial? I don’t think it is explicitly stated.

> *Indeed, the model is space-time, with a lagged dependent variable for space, and temporal correlations in the residuals. We have emphasized this in the methods.*

The methods section is well written and model structure is clearly articulated.  Although spatial SUR is not the only approach to answer your research questions, it is a very robust statistical approach.  I am curious how the results from the SUR approach compares to Geographic and Temporally Weighted Regression and Space-Time Conditional Autoregressive Models (MCMC or INLA).

> *We now comment on other potential modelling approaches in the conclusions.*

Analysis

Section 4.1 is well-written.

Figure 1 needs to be improved.  The class breaks in the legends are inconsistent, making it difficult to compare incidence rates.  The projection seems to be a bit odd.  Mainland Spain should be much larger, while the Canary Islands should be in separate insets, or remove altogether.  Finally, the black line for the Madrid label masks many provinces.

> *We changed the scale for the colors to be consistent for all maps. We tested the insets for the Canary Islands, but they crowded the figures and added relatively little information. We have decided instead to exclude them from the maps. Once this is done, the line for Madrid is less obtrusive, and overall the figures are easier to read.*

Suggest caption for Figure 2 should be “Autonomous Communities in Spain (sorted from north to south)”

> *Changed.*

I have the same suggestions for Figure 3 as I did for Figure 1.  The resolution is poor, and the mainland should be much larger and the center of the map frames.

> *Done.*

Figure 4: Are the correlations statistically significant? Please provide p-values.

>*The purpose of the daily correlation analysis in Figure 4 is to see how much day to day variability there is in the statistical associations with incidence. The p-values of of these correlation coefficients are somewhat misleading, because of the smaller samples used in their calculation. The information from this Figure helped us decide which variables to constrain in the SUR systems. At that point, the significance of the correlation is given by the p-values of the coefficients of the model. Instead of reporting the p-values for the daily correlations, we added a sentence to the effect that overall for the period  under examination, the pairwise correlations  between the independent variables and incidence are significant at $p<0.05$, with the exception of the three $Sunshine_hours$ variables. More details concerning the correlation tests can be found in the R markdown document in the chunk `correlation-tests`.*

What is the justification of using Rook instead of Queen?  Queen would strengthen the examination of spatial interaction and dependence.  I think you are losing important information with a Rook weight matrix.

> *For this revision, we used the rook criterion, and the results do not change in any major ways.*

Is it justifiable to assume that the islands are adjacent? I think that introduces a degree of uncertainty, especially after the lockdown order.

> *This is a good point. We repeated the analysis, and removing the three islands in the sample (Baleares, Las Palmas, and Tenerife) does not change the results in any major way. The results appear to be relatively robust to these changes in W. Conceptually, we agree that it makes more sense to exclude the islands given the lockdown implications. New footnotes comment on this, and the alternative analyses can be consulted in the source R markdown document.*

Figure 5: Please increase the font size of the pooled-R2 labels.

> *Done.*

Results and Discussion

I suggest that this section should just be “Discussion”, while “Analysis” should be renamed to “Results”.

> *Done.*

Page 20, Line 32: “pandemia” is misspelled.  Should be “pandemic”.

> *Typo fixed.*

Conclusion

Page 24, Lines 43-44: “Thirdly, all environmental data are based on a single station in a province”.  This is a bit concerning since a province is essentially similar to a state.  I suppose the justification that you selected a single station based on population distribution is fair, but an interpolation could have improved the accuracy and precision of the environmental variables.

> *Many thanks for this comment. We were also curious about using a summary of several stations for the environmental variables. To test this, we interpolated the environmental variables using a k-points algorithm. When using these variables, instead of seeing improvements on accuracy and precision, the goodness-of-fit of the models declines, the intervals of confidence become wider, and consequently several coefficients lose significance. It seems to us that the interpolated variables are more noisy than the manually selected stations. It is possible that a different interpolation algorithm might improve this, but that would involve experimenting with several interpolation methods to derive several environmental variables for a period of 51 days. For verification purposes, we have recreated the data package to include both datasets (single station and average of stations). In the paper we discuss only the results of the first.*

I also think that you should not finish your paper with limitations.  I suggest moving the limitations to the discussion, then finish with a strong conclusion regarding the main findings and public health impacts.

> *Thank you for this suggestion, which we have adopted.*

\newpage

## Reviewer 2

Comments to the Author
This is a well written paper exploring environmental impacts on incidence of CV19 with spatial models, controlling for a number of socioeconomic variables. There is value in this paper, but a number of issues need to be considered:

> *Thank you for your positive comments and feedback*

It would be helpful for the authors choose a name for the virus (e.g. SARS-COV2, COVID19) and use it consistently or explain if there are differences in when/how these names are use. Additionally, make sure presentation is consistent: e.g. sars-cov2 vs. sars-cov-2, COVID/Covid, etc.

> *SARS-CoV2 is the virus, and COVID-19 the disease it causes. We have clarified this, and revised the text to make consistent use of these terms.*

pg 5, ln 13 - percentage*

> *Typo was corrected.*

pg 6, ln 46 - I'm not sure I fully understand the weighting ... given the results indicate this approach doesn't produce a good fit, perhaps it can left out

> *We feel that this is, although a minor point, of interest for modellers and would prefer to retain it. We have revised the discussion of the weights and hopefully they are more clear now.*

pg 8 ln 15 "variable" misspelled, ln 16 "restrictions" misspelled.

> *Typos were corrected*

use of mean temp/humidity - does daily variation matter? For example do some regions have lows near 2 and highs near 20 c, where others have lows near 5 and highs near 6?

> *Indeed, which is why there are experiments with the moving averages of the environmental variables: 1) to smooth the daily variation in accordance to the incubation period of the disease. The narrowest window is 8 days, and the two 11-day windows weight the daily variations in different ways: uniformly for the $lagw$ variables, and giving greater weight to days closer to the median incubation period of 4-5 days in the $lag11w$ variables.*

pg 10 ln 50 - sentence grammar

could W be defined using commuter flow to gain a better sense of interaction? Are there past examples in the literature that define W for provinces across Spain? Perhaps there's more interaction between Madrid and Barcelona, etc.

> *In principle, it is possible, but unfortunately commuter flow data are not readily available. The Ministry of Transportation is working on flow data for Spain based on mobile data, to be released at a future (and yet undefined) date. In our view, once in lockdown, hierarchical flows are perhaps less relevant than cross-boundary flows, and even those should become relatively less important as the lockdown persists: this is, we believe, what the autocorrelation coefficient is telling us. See this paragraph at the end of the discussion:*

>>*Finally, we defined neighborhoods based on adjacency. It would be interesting to compare other connectivity criteria, for instance based on domestic transportation instrastructure and services. We flag this as a matter for future research.*

pg 15 ln 22 - 'transit' systems

> *Typo corrected.*

given multiple equations would it be appropriate to account for multiple testing and adjust accordingly?

> *Not with SUR, because the multiple equations are estimated jointly, and the calculations account for the size of the sample.*

pg 18 ln 32-33, do the authors mean 'due to *violations* of the restrictions of movement'?

> *This sentence was revised to conform to the slight changes in the results after the models were reestimated without the islands.*

pg 19, not sure the anecdote is appropriate - suggest this be deleted

> *Deleted.*

pg 21, ln 52 - can this recommendation be backed up with further evidence about lockdown effectiveness?

> *We think that this is a reference to pg 20, ln 52. This paragraph was too wordy, and we have edited it for clarity. Instead of effectiveness of the lockdown, we tie this discussion to risk compensation, which seems to a clearer case.*

pg 22 ln 36 - I'm not sure I follow this statement ("Without a lockdown..."). Also, please remove "!". I've tried to work through this logic, but can't quite seem to figure out how you have come to this conclusion. Wouldn't cases generated elsewhere also be affected by temperatures in those locations? Would this not make the process more complex/less clear?

> *You are right that the process becomes more complicated, and explaining it is also not simple. We were not as clear as we had hoped, but made an effort with this new paragraph:*

>>*What do these effects mean? Under a situation of lockdown, inter-regional contagion is reduced, as expected, and the total effects of the variables tend towards their direct effects. In those first few days covered by our analysis the total effect of all variables is greater due to the spatial contagion effect. This also makes analysis and intervention more complex: the contagion effect essentially acts like a multiplier, whereby developments in each province spill over to their neighbors. Once the contagion effect has been tamed, each province can be "treated" independently from its neighbors.*

please double check spelling/grammar throughout.

> *We have proof-read the paper and hopefully fixed all remaining spelling/grammar mistakes.*

\newpage

## Reviewer 3

Comments to the Author

The paper is well written and very timely. I truly enjoyed reading it and I think it should be granted publication only with some very minor revisions. Although I understand and I use econometric and statistical models, I am not an expert on the technical part and therefore I will not comment on that specifically. My comments are more "general".

> *Many thanks for your positive assessment and suggestions for revisions.*

There is a clear need of looking at environmental factors such as those you are considering. However, there has been a huge debate on pollution as a possible environmental factor affecting the spread of the disease. I would suggest that you clarify somewhere that you look only at weather-related factor and not pollution and explain why (bad data I assume?).

> *Indeed, working with meterological stations was already challenging, and we are not aware of pollution data at the scale of the analysis. We now discuss the role of pollution as an environmental factor in need of further attention.*

Working on the topic myself I often wonder how the results would change if the disease had hit somewhere else first? Clearly that would have affected the spread...Could this be a problem in your case?

>*This is an interesting question and a problem insofar as we lack a counterfactual, This is where the use of simulation techniques are useful, and the work of Muller et al. with MATSim is relevant in this respect; see:*

>>*Muller, S. A., Balmer, M., Neumann, A., & Nagel, K. (2020). Mobility traces and spreading of COVID-19. medRxiv, 2020.2003.2027.20045302. doi:10.1101/2020.03.27.20045302*

Very minor:

I would write NUTS2, NUTS3...I have almost never seen NUTII and NUTIII...

>*We made this change.*

Page 6: when you say you use a rather large geographical unit of analysis. I get that, to be quick, you have to use whatever data are available in the short run. However, I do not get what you mean by 'the analysis must therefore be considered "ecological"'. Either you explain it better or you simply remove the sentence acknowledging your results are preliminary, but could be easily applied to a finer scale once, and if, better data become available.

>*We agree, and removed the sentence to avoid overcomplicating the discussion. Earlier on in the paper we already had explained that this is a population health approach.*

Here and there I would remove some "however".

>*Done.*

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
First round of revisions after invitation to revise and resubmit.

```{r load-data, include=FALSE, cache=TRUE}
# Load data from package `covid19env`
data("covid19_spain")
data("provinces_spain")
```

@@

# Join provincial data to incidence data and convert to simple features:
covid19_spain <- covid19_spain %>% 
  left_join(provinces_spain %>% st_drop_geometry(),
            by = c("province", "CCAA", "ID_INE"))
```

@@

From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus has threatened to overrun health systems the world over. In efforts to contain the spread, numerous governments in many nations and regions have either recommended or mandated social distancing measures, and as of this writing, millions of people in five continents shelter in place. There are encouraging signs that these measures have arrested the spread of the virus where they have been implemented, and have thus helped to keep a bad situation from becoming even worse [e.g., -@Lancastle2020impact]. However, this has come at a high cost, and the consequences for all spheres of the economy, social, and cultural life have been dire [e.g., @Fernandes2020economic; @Luo2020how]. As a result, there is a sense of urgency to anticipate the progression of the pandemic, in order to plan for progressive lifting of restrictions to movement and social contact [e.g., @Kissler2020projecting]. Needless to say, this has become a delicate, and politically charged, balancing act between public health and the economy [@Gong2020balance].

A salient question in the debate on how and when to ease social distancing measures is whether the virus will display seasonal variations. Earlier, diverse studies have shown the effect of temperature and humidity on the incidence of influenza [e.g., @Makinen2009cold; @Jaakkola2014decline; @Kudo2019low]. Jaakkola et al. [-@Jaakkola2014decline], for example, found that a decrease of temperature and absolute humidity precedes the onset of symptoms of influenza A and B viruses by 3 days in places where the temperature is low. After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV. In this way, Casanova et al. [-@Casanova2010effects] used two surrogates, namely the gastroenteritis (TGEV) and mouse hepatitis viruses (MHV), to find that virus inactivation was more rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels. In a similar vein, but working directly with SARS-CoV, Chan et al. [-@Chan2011effects] found that viability of the virus was lost at temperatures higher than 38C and relative humidity superior to 95%.

While existing research on similar pathogens suggests that SARS-CoV is more stable and potentially easier to transmit in conditions of low temperature and low humidity, it is far from certain that this will also be the case with the novel SARS-CoV2. If such is the case, there is the possibility of easing restrictions to social contact as the weather warms; however, weeks or possibly months of costly measures could become undone if not, and the restrictions are lifted prematurely. Not surprisingly, given the stakes involved, this issue has already triggered a lively debate.

Some of what is thought about the possible seasonality of COVID-19 is based on analogies to the patterns of other known respiratory viruses. However, de Ángel Solá et al. [-@deangel2020weathering] note that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" (section 4). This urges caution when extrapolating from known viruses, and indeed, the evidence in this respect is inconclusive. At a global scale, de Ángel Solá et al. [-@deangel2020weathering] see less risk in the Caribean Basin; however, Coelho et al. [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate. Similarly, whereas Araujo and Naimi [-@Araujo2020spread] argue that spread of SARS-CoV2 will likely be constrained by climate, Harbert et al. [-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States. Yao et al. [-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19. This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections (Zittermann et al. 2016; Dąbrowska-Leonik et al. 2018; Dinlen et al. 2016; Mathyssen et al. 2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters, much of the relevant work has yet to be peer-reviewed and therefore is open to change (see for example the challenge of @Harbert2020spatial to @Araujo2020spread). In the United States, the National Academy of Sciences, Engineering, and Medicine was engaged by the Office of the Executive for guidance on this matter [see @National2020rapid]. Their conclusion summarizes the situation well (see p. 6): "Some limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV-2 pandemic unfolds could shed more light on the effects of climate on transmission."

With the above considerations in mind, our objective with this paper is to contribute to the knowledge basis regarding the spread of COVID-19 and the influence of environmental factors, particularly temperature, humidity, and sunshine. We adopt a population health approach, and report results from a spatial model of the incidence of COVID-19 in fifty provinces in Spain, one of the countries hardest hit by the pandemic. We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days. We then join this dataset with provincial-level economic and demographic information to act as controls to our key environmental variables. These data are analyzed using a spatial Seemingly Unrelated Regressions (SUR) approach, which allows us to model incidence of COVID-19 as a spatial contagion process.

@@

Please note that this paper is prepared as a reproducible research document. The source R Markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from the following anonymous Drive folder:

https://drive.google.com/open?id=1d_40N_QXo2Fl14r3T3CHs84IED-B1rV_

Context and Data {#context-and-data}
============

@@

The first reported case of COVID-19 in Spain was on January 31th, when a German tourist in the Canary Islands tested positive for the virus. However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia). In a short period of time, after this relatively slow start, the outbreak flared. By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic. This declaration marked a turning point for the public in Spain too. As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe, after Italy. In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th. As part of the state of emergency restrictions to most activities were imposed, with the exception of essential services (e.g. food, health) and some economic subsectors of industry and construction. A few days later, on March 17th, Spain closed its lands borders to allow entry only to returnee nationals and permanent residents. The lockdown was further hardened between March 30th and April 12th (including the Easter weekend of April 10th-12th) and during this period only essential activities were allowed. During this period, there was a dramatic reduction in overall mobility, both within provinces as between \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

## Selection of Variables {#variables}

The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global]. In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife. Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important. As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine. These variables have been implicated in the viability and ease of transmission of similar viruses. In addition to these variables, we also aim to use a set of controls, in the form of specific socio-economic and demographic characteristics of each province.

The percentaje of older adults (over 65) is the second variable that we consider as a control. Early evidence regarding COVID-19 suggests that the case rate mortality is higher at older ages [e.g. @Novel2020epidemiological]. However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission of the infection. The tool of choice in containing the spread of the disease has been social distancing. In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility]. In other words, many older adults are, whether by preference or otherwise, already in a form of social isolation. Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age]. Accordingly, our expectation is that provinces with higher percentages of older adults will tend to have similarly low levels of social contact, particularly since the age-structured matrix of social contact in Spain is similar to Italy [@Prem2017projecting].

The third population variable the we consider is population density, since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling]. The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g. cumulative incidence in urban areas like NYC). For this reason, we anticipate a positive relationship between population density and the incidence of the disease.

The last control variable is the presence of mass transit systems in a province. Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g. metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing. After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.

@@

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive. For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test. The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project by geovoluntarios.com\footnote{https://www.geovoluntarios.org/}, ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19}, and ESRI España. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments, as documented in Centro de Datos Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., *AEMET*, the state meteorology agency, and *MAPA*, the ministry of agriculture, fisheries, and food). The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute). Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.

>>>>

```{r load-data, include=FALSE, cache=TRUE}
# Load data from package `covid19env`
data("covid19_spain_1") # Use data sourced from one representative station for the province
#data("covid19_spain_2") # Use data with the average of all stations in provinces
data("provinces_spain")
```

@@

# Join provincial data to incidence data and convert to simple features:
covid19_spain <- covid19_spain_1 %>% 
#covid19_spain <- covid19_spain_2 %>% # data with the average of all stations in provinces   
  left_join(provinces_spain %>% st_drop_geometry(),
            by = c("province", "CCAA", "ID_INE"))
```

@@

From a small outbreak linked to a live animal market at the end of 2019 to a global pandemic in a matter of weeks, the SARS-CoV2 virus (and COVID-19, the disease it causes), has threatened to overrun health systems around the world. In efforts to contain the spread of the disease, numerous governments in many regions and nations have either recommended or mandated social distancing measures, and as of this writing, millions of people in five continents shelter in place. There are encouraging signs that these measures have mitigated the spread of the virus [e.g., @Lancastle2020impact; @Lewnard2020scientific; @Wilder2020isolation]. Even so, this has come at a high cost, and the consequences for all spheres of economic, social, and cultural life have been dire [e.g., @Fernandes2020economic; @Luo2020how]. As a result, there is a sense of urgency to anticipate the progression of the pandemic, in order to plan for progressive lifting of restrictions to movement and social contact [e.g., @Kissler2020projecting]. Needless to say, this has become a delicate, and politically charged, balancing act between public health and the economy [@Gong2020balance].

A salient question in the debate on how and when to ease social distancing measures is whether the virus will display seasonal variations. Existing research on similar pathogens suggests that the virus could be more stable and potentially easier to transmit in conditions of low temperature and low humidity. While encouraging this is encouraging, it is important to keep in mind that "not all seasonal respiratory viruses experience the same spatiotemporal patterns" [@deangel2020weathering, section 4]. This urges caution when extrapolating from known viruses. The evidence in this respect is as yet inconclusive, and although easing restrictions as the weather warms may appear tempting, doing so prematurely could well undo weeks or possibly months of costly measures.

It is not surprising, given the stakes involved, that this issue has already triggered a lively debate. The current state of knowledge is was well-summarized by the National Academy of Sciences, Engineering, and Medicine in the U.S. [see @National2020rapid]. Engaged by the Office of the Executive for guidance on this matter, this organization concludes that: "[some] limited data support a potential waning of cases in warmer and more humid seasons, yet none are without major limitations...Additional studies as the SARS-CoV2 pandemic unfolds could shed more light on the effects of climate on transmission" (p. 6). To further complicate matters, much of the relevant work has yet to be peer-reviewed (see for instance the challenge of @Harbert2020spatial to @Araujo2020spread).

With the above considerations in mind, our objective with this paper is to investigate the influence of environmental factors, concretely temperature, humidity, and sunshine, on the progression of the pandemic. We adopt a population health approach, and report results from a spatio-temporal model of the incidence of COVID-19 in fifty provinces in Spain, one of the countries hardest hit by the pandemic. We combine data on reported cases of the disease with metereological information, to create a spatio-temporal dataset covering a period of 30 days. We then join this dataset with provincial-level economic and demographic information to act as controls to our key environmental variables. These data are analyzed using a spatial Seemingly Unrelated Regressions (SUR) approach, which allows us to model incidence of COVID-19 as a contagion process.

@@

Please note that this paper is prepared as a reproducible research document. The source `R` markdown document, as well as all data and code needed to reproduce/review/extend the analysis can be obtained from a public repository\footnote{https://github.com/paezha/covid19-environmental-correlates}.

Background {#background}
============

The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global]. In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife. Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important. As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine.

Much of what is known about the potential seasonality of SARS-CoV2 is a result of research on other pathogens. Earlier, diverse studies have shown the effect of temperature and humidity on the incidence of influenza [e.g., @Makinen2009cold; @Jaakkola2014decline; @Kudo2019low]. Jaakkola et al. [-@Jaakkola2014decline], for example, found that a decrease of temperature and absolute humidity precedes the onset of symptoms of influenza A and B viruses by 3 days in places where the temperature is low. After the 2002-2004 outbreak of SARS, researchers also began to investigate the relationship between these factors and SARS-CoV2 [@Casanova2010effects; @Chan2011effects]. Casanova et al. [-@Casanova2010effects], for instance, used two surrogates (i.e., the gastroenteritis TGEV and mouse hepatitis MHV viruses), to find that virus inactivation was more rapid at temperatures of 20°C than 4°C, and at 40°C than 20°C; in terms of humidity, these researchers reported that survival of the virus was lower at moderate relative humidity levels. Chan et al. [-@Chan2011effects] also found that viability of the virus that causes SARS is also lost at higher temperatures (>38°C) and relative humidity superior to 95%.

Whether these results from laboratory experiments will hold in general remains uncertain. At a global scale, de Ángel Solá et al. [-@deangel2020weathering] see less risk in the Caribean Basin; however, Coelho et al. [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate. Similarly, whereas Araujo and Naimi [-@Araujo2020spread] argue that spread of SARS-CoV2 will likely be constrained by climate, Harbert et al. [-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States. Yao et al. [-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19. This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections (Zittermann et al. 2016; Dąbrowska-Leonik et al. 2018; Dinlen et al. 2016; Mathyssen et al. 2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).

In addition to the environmental variables above, we also consider a set of potential socio-economic and demographic confounders.

Secondly, we consider percentage of older adults (over 65) in a region. Early evidence regarding COVID-19 suggests that the case rate mortality is higher at older ages [e.g. @Novel2020epidemiological]. However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission of the infection. The tool of choice in containing the spread of the disease has been social distancing. In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility]. In other words, many older adults are, whether by preference or otherwise, already in a form of social isolation. Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age]. Accordingly, our expectation is that higher percentages of older adults will tend to have similarly low levels of social contact [the age-structured matrix of social contact in Spain is similar to Italy; see @Prem2017projecting].

Population density is also relevant since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling]. The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g. cumulative incidence in urban areas like NYC). For this reason, we anticipate a positive relationship between population density and the incidence of the disease.

The last variable that we consider as a control is the presence of mass transit systems in a province. Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g. metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing. After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.

Context and Data {#context-and-data}
===================

@@

The first reported case of COVID-19 in Spain was on January 31st, 2020, when a German tourist in the Canary Islands tested positive for the virus. However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia). In a short period of time, as testing started to ramp up, it became clear that an outbreak was flaring. By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic. This declaration marked a turning point for the public in Spain too. As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe after Italy. In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th. As part of the state of emergency restrictions to most activities were imposed, with the exception of essential services (e.g. food, health) and some economic subsectors of industry and construction. A few days later, on March 17th, Spain closed its lands borders to allow entry only to returnee nationals and permanent residents. The lockdown was further hardened between March 30th and April 12th (including the Easter weekend of April 10th-12th) and during this period only essential activities were allowed. During this period, there was a dramatic reduction in overall mobility, both within provinces as between \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

@@

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTS3 in Eurostat terminology) for the period between March 13th and April 11th, inclusive. For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test. The Spanish National Government publishes periodic updates at the regional level (NUTS2) and the information is also released at the provincial level as part of a collaborative project by geovoluntarios.com\footnote{https://www.geovoluntarios.org/}, ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19}, and ESRI España. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments, as documented in Centro de Datos Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., *AEMET*, the state meteorology agency, and *MAPA*, the ministry of agriculture, fisheries, and food). The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute). Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.

The spatial and temporal coverage of the data is as follows. Our dataset begins on March 13, which is the first date when every province had reported at least one case of COVID-19, and continues until April 11, for a period of 30 days. The unit of analysis is the province. Provinces are the equivalent of states, and are embedded in Autonomous Communities. As an example, Cataluña is an Autonomous Community and consists of four provinces, namely Barcelona, Gerona, Lerida, and Tarragona. The size of the provinces is relatively large, as seen in Table \ref{tab:descriptive-statistics}. The smallest province is $1,978.12km^2$ (this is smaller than Rhode Island in the U.S.) and the largest province is $21,767.93km^2$ (slightly smaller than New Jersey in the U.S.). While this is a relatively large degree of spatial aggregation, reporting on COVID-19 is not yet consistent at smaller geographies, or cases are not reported at that level at all.
